Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/towards-relational-model.tex
tweaks/comments to towards relational model section

These relations are then used to model term precision. For closed programs $M \ltdyn M' :
c$ where $c : A \ltdyn A'$ we have that these denote elements of the
free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model this
as a relation holding $\sem{M} (U\li\sem{c}) \sem{M'}$ where
$U\li\sem{c}$ is a relational lifting of $U\circ\li$.
%
More generally, $M$ and $M'$ can be \emph{open} terms, in which case
they denote functions, not just elements. We model that as not simply
being elements related by a relation but in a typical
logical-relations style of \emph{preserving} relatedness: if they are
passed in inputs related by the domain relations then their output is
related by the codomain relation.
%
To capture this preservation of relations property, we use the notion
of a \emph{square}, which specifies a relation between two monotone
functions. The definition of square is as follows. Let $A_i,
A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$
and $c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to
A_o$ and $g : A_i' \to A_o'$ be monotone functions. We say that $f
\ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$ with $(x, y)
\in c_i$, we have $(f(x), g(y)) \in c_o$. We visualize this situation
as follows:
%
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_i} \& {A_i'} \\
  {A_o} \& {A_o'}
  \arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["f"', from=1-1, to=2-1]
  \arrow["g", from=1-2, to=2-2]
  \arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%
Squares are then used to model the term precision ordering in that
$\Delta \vdash M \ltdyn N : c$ implies $\sem{M}
\ltsq{\sem{\Delta}}{U\li\sem{c}} \sem{N}$ where here $\sem{\Delta =
  x_1:c_1,\ldots}$ is the action of the product on the relations
$c_1,\ldots$.

@@

% \max{This needs to also talk about relational composition and horizontal pasting squares, since that's transitivity!}
We note that for any $c$, there is a ``vertical identity'' square $\id
\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
\ltsq{\le_{A_i}}{\le_{A_o}}$ by the definition of monotone function.
%
We can \emph{compose} squares vertically, i.e., if we have $f \ltsq{c_1}{c_2}
f'$ and $g \ltsq{c_2}{c_3} g'$ then we obtain a square $g \circ f
\ltsq{c_1}{c_3} g' \circ f'$. Likewise, we can compose squares horizontally: If
$f \ltsq{c_i}{c_o} g$ and $g \ltsq{c_i'}{c_o'} h$, then we obtain a square $f
\ltsq{c_i \comp c_i'}{c_o \comp c_o'} h$. Horizontal composition of squares
corresponds to transitivity of term precision.

@@

Consider the UpL rule. It states that if
$c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $M : A_1$ is
related to $N : A_3$ via the composition $c \comp c'$, then the upcast
$\upc{c} M$ is related to $N$ via $c'$. Since the upcasts are pure
functions in the semantics this can be deduced from the existence of
the following square, where the relation on the top is $\sem{c} \comp \sem{c'}$:
% \max{the square doesn't make sense because we
%   haven't discussed horizontal composition of relations or that this
%   notation means horizontal composition of relations}
%
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {\sem{A_1}} \& {\sem{A_2}} \& {\sem{A_3}} \\
  {\sem{A_2}} \&\& {\sem{A_3}}
  \arrow["\sem{c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["{\sem{\upc{c}}}"', from=1-1, to=2-1]
  \arrow["{\sem{c'}}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
  \arrow["\id", from=1-3, to=2-3]
  \arrow["{\sem{c'}}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
\end{tikzcd}\]
While at first look this seems to be specifying a relationship between
$\sem{c}$ and $\sem{\upc c}$, there is a problem: it also quantifies
involves an arbitrary other relation $\sem{c'}$! This means we cannot
require the existence of this square as part of the definition of a
relation between value types, as it is self-referential. This would
seem to imply that we cannot give a compositional model for
graduality. However, New and Licata observed that in the presence of
transitivity, we can \emph{derive} the above squares from simpler ones
that do not involve composition of relations. Below is the square
corresponding to UpL:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \\
	{A_2} \& {A_2}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["\id", from=1-2, to=2-2]
	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%
This square, with a dual one related to UpR, say that the relation $c$
is \emph{representable} by the upcast morphism $\upc{c}$, essentially
that the relation is a kind of \emph{graph} of the function in that $x
\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$.

>>>>

These relations are then used to model term precision. For closed programs $M \ltdyn M' :
c$ where $c : A \ltdyn A'$, the terms denote elements of the
free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model the term precision semantically as the relation holding between those denoted elements $\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$ where
$U\li\sem{c}$ is a relational lifting of $U\circ\li$.
%
More generally, $M$ and $M'$ can be \emph{open} terms, in which case
they denote functions, not just elements. We model that as not simply
elements being related but in a typical logical-relations style as
\emph{preserving} relatedness: if they are passed in inputs related by
the domain relations then their output is related by the codomain
relation.
%
To capture this preservation of relations property, we use the notion
of a \emph{square}, which specifies a relation between two monotone
functions. The definition of square is as follows. Let $A_i, A_o,
A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
$c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to A_o$
and $g : A_i' \to A_o'$ be monotone functions. We say that $f
\ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$ with $(x, y)
\in c_i$, we have $(f(x), g(y)) \in c_o$. We call this a \emph{square}
because we visualize this situation as follows:
%
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_i} \& {A_i'} \\
  {A_o} \& {A_o'}
  \arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["f"', from=1-1, to=2-1]
  \arrow["g", from=1-2, to=2-2]
  \arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%
Squares are then used to model the term precision ordering in that
$\Delta \vdash M \ltdyn N : c$ implies $\sem{M}
\ltsq{\sem{\Delta}}{U\li\sem{c}} \sem{N}$ where here $\sem{\Delta =
  x_1:c_1,\ldots}$ is the action of the product on the relations
$c_1,\ldots$.

@@

% \max{This needs to also talk about relational composition and horizontal pasting squares, since that's transitivity!}
We note that for any $c$, there is a ``vertical identity'' square $\id
\ltsq{c}{c} \id$, as can be seen by unfolding the
definition. Similarly, for any montone function $f : A_i \to A_o$,
there is a ``horizontal identity'' square $f
\ltsq{\le_{A_i}}{\le_{A_o}} f$ arising from the fact that $f$ is
monotone.
%
We can \emph{compose} squares vertically, i.e., if we have $f
\ltsq{c_1}{c_2} f'$ and $g \ltsq{c_2}{c_3} g'$ then we obtain a square
$g \circ f \ltsq{c_1}{c_3} g' \circ f'$. Likewise, we can compose
squares horizontally: If $f \ltsq{c_i}{c_o} g$ and $g
\ltsq{c_i'}{c_o'} h$, then we obtain a square $f \ltsq{c_i \comp
  c_i'}{c_o \comp c_o'} h$. Horizontal composition of squares
corresponds to transitivity of term precision. While we do not dwell
on categorical abstractions in this work, we note that this structure
of functions, relations and squares forms a locally thin \emph{double
category} as used in previous work \cite{newlicata}.

@@

Consider the UpL rule. It states that if
$c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $M : A_1$ is
related to $N : A_3$ via the composition $c \comp c'$, then the upcast
$\upc{c} M$ is related to $N$ via $c'$. Since the upcasts are pure
functions in the semantics this can be deduced from the existence of
the following square, where the relation on the top is $\sem{c} \comp \sem{c'}$:
% \max{the square doesn't make sense because we
%   haven't discussed horizontal composition of relations or that this
%   notation means horizontal composition of relations}
%
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {\sem{A_1}} \& {\sem{A_2}} \& {\sem{A_3}} \\
  {\sem{A_2}} \&\& {\sem{A_3}}
  \arrow["\sem{c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["{\sem{\upc{c}}}"', from=1-1, to=2-1]
  \arrow["{\sem{c'}}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
  \arrow["\id", from=1-3, to=2-3]
  \arrow["{\sem{c'}}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
\end{tikzcd}\]
While at first look this seems to be specifying a relationship between
$\sem{c}$ and $\sem{\upc c}$, there is a problem: it also quantifies
involves an arbitrary other relation $\sem{c'}$! This means we cannot
require the existence of this square as part of the definition of a
relation between value types, as it is self-referential. This would
seem to imply that we cannot give a compositional model for
graduality. However, New and Licata observed that in the presence of
transitivity, we can \emph{derive} the above squares from simpler ones
that do not involve composition of relations. Below is the square
corresponding to UpL:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \\
	{A_2} \& {A_2}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["\id", from=1-2, to=2-2]
	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%
This square, with a dual one related to UpR, say that the relation $c$
is \emph{representable} by the upcast morphism $\upc{c}$, essentially
that the relation is a kind of \emph{graph} of the function in that $x
\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$ \cite{shulmandoublecategories}.
